A Technique for Inter-Catalogue Recommendations

### 2014-06-11 paper2/paper.tex
Absorb Implementation section into Algorithms section.

A key feature of tf-idf is that terms appearing in many item descriptions are
discounted via the idf factor. Conversely, rare terms will have larger tf-idf
values. These features help capture the most relevant words in each description.
We now define the $ContentSimilarity$ between items $A$ and $B$ as
\begin{align}
ContentSimilarity(A,B) = \vec{\tau}(A) \cdot \vec{\tau}(B)
\end{align}
where $\vec{\tau}(A)$ is the tf-idf vector representation of item $A$.

@@

A common approach to learning latent features is to apply matrix factorization
techniques to user-item data, as in many CF methods \cite{Koren2009}. But,
user-item data, a prerequisite for such techniques, in not available to us here.
A key insight of the present work is that publicly available recommendation
graphs, such as those found on Macy's website, can be used to simulate user-item
data through the use of special random walks known as {\em Topic-Sensitive
PageRank}, a technique typically associated with web search
\cite{Haveliwala2002}.

@@

Constructing our random walk in this way provides an important benefit. In the
context of the random walk, the recommendation graph may be considered a {\em
discrete-time Markov chain}. By providing the walker a finite probability of
staying at each item, our chain becomes {\em aperiodic}. And since our walker
begins at $A$ and can only teleport to $A$, she is effectively confined to a
strongly connected subgraph of the recommendation graph. This subgraph is
therefore, by construction, {\em irreducible}.  Importantly, according to the
{\em Fundamental Theorem of Markov Chains}, any finite, aperiodic, irreducible
Markov chain has a unique stationary distribution \cite{Norris1998}.

We may then construct a synthetic $N \times N$ user-item matrix $M$ with rows
representing items and columns representing users, which are in fact the
stationary distributions discovered from our random walks. This matrix has at
least one notable advantage over most real user-item matrices---it is not
sparse.  Rather, each synthetic user has a finite value associated with every
item that is reachable from its origin item in the graph. Unlike the typical CF
scenario, we may therefore employ SVD as our matrix factorization technique
since our user-item matrix is virtually complete (i.e. nearly void of empty
entries).  Using SVD, we factor our user-item matrix as
\begin{align}
M = U \Sigma V^T,
\end{align}
where the rows of the $N \times N$ matrix $U$ and the $N \times N$ matrix $V$
are the items and users represented in an $N$-dimensional concept space,
respectively, and the $N$ diagonal entries of $\Sigma$ are the singular values
associated with each concept.

Having transformed our items into concept space, we now discard all but the $k$
leftmost columns of the matrix $U$. The idea is that these leftmost columns are
the most significant eigenvectors of concept space and capture most, if not all,
of the relevant latent features. Meanwhile, the discarded columns are presumed
to contain mostly noise. This rank-reduction process yields matrices that are
typically written as
\begin{align}
M \approx U_k \Sigma_k V^T_k,
\end{align}
where the rows of $U_k$ and $V_k$ are the items and users represented in a
$k$-dimensional concept space, respectively.  Rank-reducing following the
application of SVD in this way is a common technique used in information
retrieval and is often referred to as {\em Latent Semantec Indexing} (LSI)
\cite{Deerwester1990}.

Using random walks to simulate user data, followed by the application of LSI, is
indeed an effective way to encode information from a recommendation graph into a
latent feature space. But what is the relationship between latent feature
vectors derived from disconnected graph partitions? In principle, there is no
guarantee of any relationship between such concept spaces in the same way that
there is no guarantee of any relationship between two arbitrary, disconnected
recommendation graphs. Nonetheless, our hypothesis is that a relationship
between these concept spaces often will exist due to the effectiveness of the
underlying CF techniques that generate the graphs to capture meaningful
item-item associations. Put another way, we believe that the most fundamental
latent dimensions along which products in a category such as Women's swimsuits
vary will be common to many swimsuit catalogues. If this is indeed true, then it
might be possible to learn a mapping between these concept spaces.

If a mapping between the concept spaces associated with two catalogues were
known, we could translate all items into a common space and then choose
inter-catalogue edges between item pairs that are nearest to one another in
concept space. To learn these mapping, we must revisit the notion of content
similarity. Once again, we utilize tf-idf vectors. This time, however, we weight
an item $i$'s tf-idf vector $\vec{\tau}(i)$ by the elements in its latent
feature vector $\vec{v}(i)$. In this way, we construct a dense {\em
super-tf-idf} vector for each concept for each catalogue. Formally, we say that
a concept $j$ of a catalogue $C$ is represented by the super-tf-idf vector
\begin{align}
\vec{s}(j,C) = \sum_{i \in C}{v(i)_j \cdot \vec{\tau}(i)},
\end{align}
where $v(i)_j$ is the $j$\textsuperscript{th} component of item $i$'s latent
feature vector. These super-tf-idf vectors act as a proxy for the concept. We
approximate the mapping between two concept spaces by taking inner products of
super-tf-idf vectors. The details of this mapping process are provided in a
subsequent section.

@@

%Now for a concept such as {\em formalness}, the words ``sophisticated'' or
%``classy'' may have high value since products with those descriptive words would
%also have a large amount of the {\em formalness} concept in their latent feature
%vectors.

\section*{Implementation}
Let us dig deeper into the details of the LSI approach described above. There 
are a number of implementation decisions that, in all, dramatically affect the
performance of this method.

\subsection*{Random Walks}
The first step of the LSI approach is to simulate user sessions by randomly
walking the recommendation graphs. There are a few ways this could be
implemented.  One method is to randomly assign a random walker a starting node
and then allow it to randomly walk the edges, with a probability of termination
at each step.  Running this many times is a straight-forward model of user
sessions.

To speed up the simulation, we calculate a series of product-specific stationary
distributions. As with any random walk technique, we first interpret the
recommendation graph as a transition matrix. In this walk, for each product we
create a product-specific transition matrix which has added teleportation edges
back the that product. We calculate the stationary distributions of each of
these matrices, one per product. We use these product-specific stationary
distributions as our user sessions. Since there is only one user session per
product, we have many fewer sessions than we would with a real dataset. However,
each of these stationary distributions can be interpreted as an aggregation of
an infinite number of random walkers starting at each node and travelling until
a random termination time, as in the original model.

This process requires a couple of parameters. The first is the number of steps
used to approximate each product-specific statinary distribution. The second
parameter is the likelihood of teleporting to the special product. For our
purposes we use 30 steps and a teleportation probability of 5\%.

\subsection*{Singular Value Decomposition}
Unlike Principal Compenent Analysis, standard LSI does not include any
preprocessing of the data, such as subtracting the mean from each column of the
product-session matrix. A major reason for this is that LSI is typically applied
to natural language processing problems, where the rows are terms, and there is
one term per word in the vocabulary. This otherwise sparse matrix becomes dense
if the mean is subtracted, making many NLP datasets untractable to work with. 
Our datast, on the other hand, has significantly fewer rows and columns and is
already dense. Therefore, there is little added cost to subtracting the
means. We include this preprocessing step and find that we get better results
because of it.

Another design decision is in the size of the truncation of the matrices learned
from SVD. This number dictates the number of latent concepts learned for LSI.
Too few concepts fail to capture enough detail to make accurate representations
of each product. Too many make mapping between concepts challenging. We found a
good balance at 16 concepts.

[Graphs of concepts]

\subsection*{Concept Mapping}
A tricky operation of this procedure is the concept mapping. We attempt to find
the best linear transformation from one concept space to the other. We do this
by representing each concept as a vector of words in the vocabulary, each
weighted with how well it represents the concept. We construct this by weighing
the tf-idf of each word in a product's description by how well that word
represents the concept. Note that products can have a positive or negative value
for each concept, thus the word representation of concpets are also positive and
negative.

With these concept representations, we construct an innner-product matrix. The
value of a cell (i, j) in the matrix is the dot product of concept i in the
first space with concept j in the second. The columns and rows of the resulting
matrix represent the relative similarity of the descriptive words between two
concepts.  But the scale of this matrix can be very large. If the inner-product
matrix is used without any normalization as the linear transformation between
the concept spaces, the scale of the transformed products will be inconsistent
with those originally in that space.

One way to normalize is to normalize each column of the inner-product matrix by
the L1 norm. This preserves the L1 norm of products after the transformation,
but does not preserve the L2 norm. Similarly, we can normalize the columns by
the L2 norm. Neither of these methods preserve the L2 norm, which would be ideal
as we find the nearest neighbors in the other concept space by their L2
distance.

Another approach is to orthogonalize the inner-product matrix. That is, find the
orthogonal matrix nearest to the inner-product matrix. Nearness is defined by
the NORM FUNCTION on the difference of the two matrices. This orthogonal matrix
does perserve the L2 norm of the vectors it transforms, however the
transformation is an approximation of the ideal inner-product matrix and can
distort the relative mapping from one concept to the others.

A third approach to normalizing aims to preserve the angle of the inner-product
transformation. For this approach, we remove the issue of scale by normalizing
all products in the mapped-to space to be unit lenght. Then, after transforming
the products from the other catalogue, we normmalize them as well. Thus the
scales for both sets of products are the same and the angles of the
inner-product transformation are preserved, at the expense of discarding the
norms of all products in their product spaces. This is equivalent to calculating
the cosine similarity between products without any normalization, as cosine
similarity is indeifferent to scale. We find that this method of normalization
performs the best.

Further gains can be seen if we consider bigrams in the description as well as
the unigrams. Here, we treat bigrams as a separate set dimensions to represent
each concept. These are included in the inner product between two concepts,
weighted compared to the unigrams by a parameter $\beta$. $\beta = 0.3$ improves the
performance of the concept mapping.

\subsection*{Popularity}
The most effective addition to the LSI procedure is incorporating the popularity
of products. We do not know a priori how popular an item is as we do not have
explicit user data. However, we can use the recommendation graphs to model the
popularity of each item. The assumption is that a retailer will recommend
popular products more frequently than unpopular ones.

We imagine that recommendation graphs are constructed as follows: A retailer first
builds a similarity matrix, using a collaborative filtering method. In this
matrix, all similarity relationships are symmetric. Next, the retailer modifies
the values of the matrix to include popularity. As a simple model, we say that
the popularity-modified similarity score from Product i to Proudct j is the
simple similarity score at (i, j) multiplied by the popularity of Product j. In
this way, more popular items are more frequently recommended.

We see signs of this in the undirected recommendation graphs. Although each product
has at most four outgoing recommendations on macys.com, we see that some
products have over one hundred incoming recommendations. With this in mind, we
approximate the popularity of each product as the logarithm of the degree of
that product in its undirected recommendation graph. We choose to model it as the
logarithm because a few of the products tend to have degrees in the hundreds.
Without some way of scaling back these popularities, these few very popular
products tend to take over all recommendations.

[Graph of popularity]

Assuming this model of recommendation graph construction, which separates similarity
from popularity, we seek to isolate similarity from the recommendation graph. Since
popular products were promoted for popularity, we demote them when walking the
graphs to create user sessions. In practice, we divide the likelihood of walking
an edge by the popularity of the product on the other side of that edge. Thus
the concepts learned on that matrix better represent products independent of
popularity.

Proceeding as before, we find the nearest products in concept space. However,
before deeming the nearest neighbors to be the best recommendations, we divide
the distance to each product by its popularity. Since we model distance in 
concept space to be inversely proportional to similarity, this reincorporation
of popularity is analagous to multiplying similarity by popularity, as done in
the original construction of the recommendation graph.

A final way in which we incorporate popularity is in the number of
recommendations alotted per product. Since we model the recommendation graph as an
undirected graph, different nodes have different degrees. A product that is
popular within its catalogue is likely to be more popular across catalogues as
well. Therefore, we allow popular products to have more recommendations across
the two catalogues, proportional to their popularity.

In all, these popularity additions significantly boost performance in the
experiment. The results of this, along with the other modifications outlined in
this section, can be seen in the following section on performance.

>>>>

A key feature of tf-idf is that terms appearing in many item descriptions are
discounted via the idf factor. Conversely, rare terms will have larger tf-idf
values. These features help capture the most relevant words in each description.

We can further improve our content similarity function by considering the
bigrams that appear in item descriptions. To exploit this fact, we construct
another tf-idf vector for each item, this one corresponding to the observed
bigrams. Combining these tf-idf vectors, we define the $ContentSimilarity$
between items $A$ and $B$ as
\begin{align}
ContentSimilarity(A,B) = \vec{\tau}_1(A) \cdot \vec{\tau}_1(B) + \beta \cdot
\vec{\tau}_2(A) \cdot \vec{\tau}_2(B),
\end{align}
where $\vec{\tau}_1(A)$ is the unigram tf-idf vector and $\vec{\tau}_2(A)$ is
the bigram tf-idf vector representations of item $A$, and $\beta$ is a
parameter. We find that that $\beta=0.3$ works well for our data.

@@

\subsubsection*{Simulating User Data}
A common approach to learning latent features is to apply matrix factorization
techniques to user-item data, as in many CF methods \cite{Koren2009}. But,
user-item data, a prerequisite for such techniques, in not available to us here.
A key insight of the present work is that publicly available recommendation
graphs, such as those found on Macy's website, can be used to simulate user-item
data through the use of special random walks known as {\em Topic-Sensitive
PageRank}, a technique typically associated with web search
\cite{Haveliwala2002}.

@@

Constructing our random walk in this way provides an important benefit. In the
context of the random walk, the recommendation graph may be considered a {\em
discrete-time Markov chain}. By providing the walker a finite probability of
staying at each item, our chain becomes {\em aperiodic}. And since our walker
begins at $A$ and can only teleport to $A$, she is effectively confined to a
strongly connected subgraph of the recommendation graph. This subgraph is
therefore, by construction, {\em irreducible}.  Importantly, according to the
{\em Fundamental Theorem of Markov Chains}, any finite, aperiodic, irreducible
Markov chain has a unique stationary distribution \cite{Norris1998}. For our
data, we find that 30 steps with a 5\% probability of teleporting back to the
origin item yields a good approximation of the true stationary distribution. We
then interpret the distribution associated with each item as the preference data
assoicated with a particular user.

\subsubsection*{Extracting Latent Features}
We may now construct a synthetic $N \times N$ user-item matrix $M$ with rows
representing items and columns representing users, which are in fact the
stationary distributions discovered from our random walks. This matrix has at
least one notable advantage over most real user-item matrices---it is not
sparse.  Rather, each synthetic user has a finite value associated with every
item that is reachable from its origin item in the graph. Unlike the typical CF
scenario, we may therefore employ SVD as our matrix factorization technique
since our user-item matrix is virtually complete (i.e. nearly void of empty
entries).  Using SVD, we factor our user-item matrix as
\begin{align}
M = U \Sigma V^T,
\end{align}
where the rows of the $N \times N$ matrix $U$ and the $N \times N$ matrix $V$
are the items and users represented in an $N$-dimensional concept space,
respectively, and the $N$ diagonal entries of $\Sigma$ are the singular values
associated with each concept.

Having transformed our items into concept space, we now discard all but the $k$
leftmost columns of the matrix $U$. The idea is that these leftmost columns are
the most significant eigenvectors of concept space and capture most, if not all,
of the relevant latent features. Meanwhile, the discarded columns are presumed
to contain mostly noise. This rank-reduction process yields matrices that are
typically written as
\begin{align}
M \approx U_k \Sigma_k V^T_k,
\end{align}
where the rows of $U_k$ and $V_k$ are the items and users represented in a
$k$-dimensional concept space, respectively. For our data, we find that $k=16$
yields good results across all categories, striking a balance between
information loss and the tractability of mapping concept spaces (see below).
Rank-reducing following the application of SVD in this way is a common technique
used in information retrieval and is often referred to as {\em Latent Semantec
Indexing} (LSI) \cite{Deerwester1990}.

\subsubsection*{Concept Space Mapping}
Using random walks to simulate user data, followed by the application of LSI, is
indeed an effective way to encode information from a recommendation graph into a
latent feature space. But what is the relationship between latent feature
vectors derived from disconnected graph partitions? In principle, there is no
guarantee of any relationship between such concept spaces in the same way that
there is no guarantee of any relationship between two arbitrary, disconnected
recommendation graphs. Nonetheless, our hypothesis is that a relationship
between these concept spaces often will exist due to the effectiveness of the
underlying CF techniques that generate the graphs to capture meaningful
item-item associations. Put another way, we believe that the most fundamental
latent dimensions along which products in a category such as Women's swimsuits
vary will be common to many swimsuit catalogues. If this is indeed true, then it
might be possible to learn a mapping between these concept spaces.

If a mapping between the concept spaces associated with two catalogues were
known, we could translate all items into a common space and then choose
inter-catalogue edges between item pairs that are nearest to one another in
concept space. To learn these mapping, we must revisit the notion of content
similarity. Once again, we utilize tf-idf vectors. This time, however, we weight
an item $i$'s tf-idf vector $\vec{\tau}(i)$ by the elements in its latent
feature vector $\vec{v}(i)$. In this way, we construct a dense {\em
super-tf-idf} vector for each concept for each catalogue. Formally, we say that
a concept $j$ of a catalogue $C$ is represented by the super-tf-idf vector
\begin{align}
\vec{s}(j,C) = \sum_{i \in C}{v(i)_j \cdot \vec{\tau}(i)},
\end{align}
where $v(i)_j$ is the $j$\textsuperscript{th} component of item $i$'s latent
feature vector. Note that since the components $v(i)_j$ may be positive or
negative so too can the elements of a concept's super-tf-idf vector, in
constrast to the elements of an individual item's tf-idf vector which are
necessarily greater than or equal to zero. Also note that we compute both
$\vec{s}_1$ and $\vec{s}_2$, which correpond to unigram and bigram super-tf-idf
vectors, respectively, for each concept. These super-tf-idf vectors then act as
proxies for the concept.

With these proxy representations of the concepts for the concept spaces of two
catalogues $C_1$ and $C_2$, we construct a matrix $A$ of inner-products of
super-tf-idf vectors.  Specifically, the matrix elements of $A$ are given by
\begin{align}
A_{ij}(C_1,C_2) = \vec{s}_1(i,C_1) \cdot \vec{s}_1(j,C_2) + \beta \cdot
\vec{s}_2(i,C_1) \cdot \vec{s}_2(j,C_2),
\end{align}
where $i$ is a concept from $C_1$ and $j$ is a concept from $C_2$ and $\beta$ is
the same parameters as before (again, we find $\beta=0.3$ works well).  We may
now use this matrix to transform the vector representation of an item in one
concept space into a vector representation in another concept space.

It is important to note that the transformation matrix $A$ is not orthogonal, and
therefore, does not preserve vector norms. We address this issue by normalizing
all feature vectors, both foreign and domestic, in the concept space following
the transformation. This radially projects all items onto the surface of sphere
within the 16-dimensional concept space. Now when ranking an item's neighbors by
distance, we are actually ranking neighbors by angular separation. That is, we
are ranking neighbors by their cosine similarity in concept space. After
experimenting with various normalization schemes we find that this approach
works best.

@@

\subsubsection*{Popularity}
The $LatentFeatureRecommender$ can be significantly improved by incorporating
the notion of item {\em popularity}. The idea is that a more popular item is
more likely to appear as a recommendation independent of how similar it is.  We
see signs of this in the undirected recommendation graphs from www.macys.com.
Although each product has at most four outgoing recommendations on its details
page, we see that some products have over one hundred incoming recommendations.
But how might we define {\em popularity}? A reasonable definition would
introduce a quantity that is correlated with the number of users that indicate a
preference for an item. Like all user data, this is not publicly available
information.  Once again, however, we can leverage the recommendation graph to
infer an approximate popularity metric for each item.

We define item popularity as
\begin{align}
\rho(A) = \mathrm{log}(1 + \mathrm{deg}(A)),
\end{align}
where $\mathrm{deg}(A)$ is the degree of the item $A$ within its recommendation
graph.  We find this quanity to be useful in improving the
$LatentFeatureRecommender$ in several ways.

We begin with an attempt to seprate the notion of item similarity from that of
item popularity. The idea that more popular items are more likely to be
recommended than less popular items motivates an alternative mode for conducting
our random walks over a recommendation graph in order to better isolate item
similarity. We do this by adjusting the probability of transitioning to an item
on the graph by making it inversely proportional to that item's popularity
$\rho$. Stationary distributions learned from these alternative random walks
better reflect item-item similarity. We find that removing item popularity from
our simulated user data results in better success when mapping between the
derived concept spaces (as evidenced by the improved precision and recall
discussed in the next section).

Next, we reincorporate popularity by modifying how we search for nearest
neighbors in concept space. Instead of simply ranking neighbors by distance, we
now rank by distance divided by $\rho$.  In this context, having defined
popularity as the logorithm of node degree reduces the liklihood that one or two
very popular items will always appear as the best recommendations when using
this weighted ranking policy.

The third way in which popularity is used is determining the number
recommendations alotted per product. We assume that a popular product in one
catalogue is likely to be a popular product in another catalogue. Specifically,
the number of predicted inter-catalogue recommendations made per item is
made directly proportional to the item's value of $\rho$.

In all, these popularity additions significantly boost the performance of the
\linebreak $LatentFeatureRecommender$. The results of this, along with
comparisons to the other inter-catalogue  recommeder algorithms described in
this section, are provided in the following section.
